Modelling the creation of ecological traps by environmentally persistent pathogens

### 2015-06-08 Manuscript/manuscript.tex
Exploratory plots and additional edits

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

Despite facilitating pathogen spread, for pathogens with either low or high disease-induced mortalities, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (Fig~\ref{highvlow}).  
In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat.
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely regardless of habitat quality, high quality distributions again supported larger host populations than low quality distributions.

For pathogens falling between these extremes (i.e. with intermediate disease-induced mortality), the habitat quality distribution had a substantial effect on pathogen spread.  
In particular, for these pathgoens, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).  
For a given pathogen in this range (here with $\nu = 0.2$ and $\delta = 0.3$), the relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the more pronounced effect of increased longevity on pathogen persistence and spread in high quality habitat distributions (note that this is not a general trend and in some cases is even reversed).
Again, metapopulations with a lattice structure displayed similar patterns, but required lower disease-induced mortality to see the benefit of low quality habitat (Supplement).

@@

This work demonstrates that for many pathogens, the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting size and stability of the host population.  
These impacts are driven largely by the interaction between pathogen traits (i.e. disease-induced mortality, transmission rates) and the colonization and extinction rates of high and low quality habitat.

The general findings that low quality habitat can inhibit disease spread while high quality habitat can enhance it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models.
And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
Broadly, we can divide the observed dynamics into three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is a net benefit; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is a net benefit; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again a net benefit.

@@

Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Figs~\ref{endemic}, \ref{highvlow}), high quality habitat is again a net benefit to the total host population size.  
Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
Here then, the benefit of high quality habitat relative to low quality derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.

@@

However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e. at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
In particular, the ability of these pathogens to persist and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play.

@@

Thus, if high quality habitat is preferentially selected by migrants (i.e. $\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
This fitness cost (i.e. the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat.
Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

The relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
Indeed, for pathogens with traits in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preferential selection of high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists.
These infectious colonists then help to maintain the environmental reservoir and the trap effect.  
Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. low population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.  
As a result, low quality patches effectively represent a dead-end the pathogen, reducing the number of patches from which it can spread, and thereby reducing total infectious occupancy and maintaining larger host population sizes (Fig~\ref{endemic}, \ref{popsizes}).

Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall metapopulation stability.
Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
Indeed, we see that the combination of the trap effect (i.e. host preference for high quality habitat) along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction (Fig~\ref{pext}).
From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}). 
Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Similarly, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.

@@

This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.  
In cases where the host movement rate was either low enough that the pathogen was unable to persist or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate (i.e. endemic) pathogen spread, factors that increase host metapopulation stability and connectivity (e.g. increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.

However, we find that as the environmental longevity of the pathogen increases, the region in which we observe these effects becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a. to c.).
These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. when the pathogen is long-lived in the environment), factors that facilitate metapopulation stability (i.e. increased movement, high quality patches) are generally a greater benefit than factors that inhibit pathogen spread (i.e. decreased movement, low quality patches).

>>>>

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

Despite generally facilitating pathogen spread, for a range of pathogens, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (red shaded regions in Fig~\ref{highvlow}).  
In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left of panels in Fig~\ref{highvlow}).
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e. at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).

However, for pathogens falling between these extremes (e.g. with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, the more widespread disease-induced mortality faced by hosts on high quality habitat distributions results in smaller total population sizes than on low quality habitat.
As pathogen longevity increases, this range of pathogens for which low quality habitat is beneficial to the host population becomes smaller, with this benefit apparent only for the most virulent pathogens.
Again, metapopulations with a lattice structure displayed similar patterns (Supplement).

@@

This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.  
These impacts are driven largely by the interaction between pathogen traits (i.e. disease-induced mortality, environmental and direct transmission rates) and the colonization and extinction rates of high and low quality habitat.

The findings that low quality habitat can inhibit disease spread while high quality habitat can enhance it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
In this light, this work expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models.
And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
Broadly, we can divide the observed dynamics into three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is a net benefit; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is a net benefit; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again a net benefit.

@@

Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Figs~\ref{endemic}, \ref{highvlow}), high quality habitat is again a net benefit to the total host population size.  
Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
Here then, the benefit of high quality habitat relative to low quality derives primarily from its ability to better maintain metapopulation stability in the face of widespread disease-induced mortality.

@@

However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e. at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
In particular, the ability of these pathogens to persist and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.

@@

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
This fitness cost (i.e. the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat.
Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

The relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
Indeed, for pathogens with traits in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists.
These infectious colonists then help to maintain the environmental reservoir and the trap effect.  
Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality habitat facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. low population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.  
As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection.
By preventing the widespread disease-induced mortality observed in high quality habitat distributions, low quality distributions are then able to maintain larger host population sizes (Fig~\ref{endemic}, \ref{popsizes}).

Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
Indeed, we see that the combination of the trap effect (i.e. host preference for high quality habitat) along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction (Fig~\ref{pext}).
From a management perspective, it is critical to understand this shift in the effect of habitat quality, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}). 
Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Similarly, our model suggests that in some cases, low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.

@@

This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.  
In cases where the host movement rate was either low enough that the pathogen was unable to persist or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate (i.e. endemic) pathogen spread, factors that would normally increase host stability and connectivity in the absence of disease (e.g. increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.

However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a. to c.).
Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).
